---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## Loading and preprocessing the data

```{r, echo=TRUE}
unzip(zipfile = 'activity.zip')
activity_data <- read.csv(file="activity.csv", header=T)
```


## What is mean total number of steps taken per day?

```{r, echo=TRUE}
library(ggplot2)
library(plyr)

activity_data_without_na <- activity_data[complete.cases(activity_data),]

activity_by_day <- ddply(activity_data_without_na, "date", summarize, total_steps = sum(steps))

ggplot(data=activity_by_day, aes(total_steps)) + geom_histogram(binwidth=1000) + 
    labs(x="Total number of steps taken each day")
```

```{r, echo=TRUE}
mean_total_steps_by_day <-  mean(activity_by_day$total_steps, na.rm=T)
mean_total_steps_by_day
```

```{r, echo=TRUE}
median_total_steps_by_day <-  median(activity_by_day$total_steps, na.rm=T)
median_total_steps_by_day
```

The mean total number of steps taken each day is `r round(mean_total_steps_by_day,0)`.  

The median total number of steps taken each day is `r round(median_total_steps_by_day,0)`.


## What is the average daily activity pattern?

```{r, echo=TRUE}
activity_by_interval <- ddply(activity_data, "interval", summarize, average_steps = mean(steps, na.rm=T))

ggplot(data=activity_by_interval, aes(x=interval, y=average_steps)) + geom_line() +
    labs(x="5-minute interval", y="Average number of steps taken") +
    labs(title="Average number of steps taken for each 5-minute interval across all days") 
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, echo=TRUE}
interval_with_max_steps <- activity_by_interval[which.max(activity_by_interval$average_steps),]
interval_with_max_steps
```

The `r interval_with_max_steps$interval` 5-minute interval contains the maximum number of steps of `r round(interval_with_max_steps$average_steps,0)` steps.


## Imputing missing values

```{r, echo=TRUE, results="hide"}
sum(is.na(activity_data$steps))
```

Based on the dataset, there are `r sum(is.na(activity_data$steps))` rows with missing steps values.

Fill in all the missing values with the mean for its 5-minute interval

```{r, echo=TRUE}
activity_data_merge = merge(activity_data, activity_by_interval, by = "interval")

activity_data_merge$steps_modified = ifelse(is.na(activity_data_merge$steps),activity_data_merge$average_steps, activity_data_merge$steps) 
```

Let's see the sample of the new dataset

```{r, echo=TRUE, results="hide",message=FALSE}
library(dplyr)
```

```{r, echo=TRUE}
set.seed(10)
sample_n(activity_data_merge, 10)
```

Let's look at the histrogram of the new dataset

```{r, echo=TRUE}
activity_data_merge_by_day <- ddply(activity_data_merge, "date", summarize, total_steps = sum(steps_modified, na.rm=T))

ggplot(data=activity_data_merge_by_day, aes(total_steps)) + geom_histogram(binwidth=1000) + 
    labs(x="Total number of steps taken each day")
```

```{r, echo=TRUE}
mean_by_day <-  mean(activity_data_merge_by_day$total_steps, na.rm=T)
mean_by_day
```

```{r, echo=TRUE}
median_by_day <-  median(activity_data_merge_by_day$total_steps, na.rm=T)
median_by_day
```

For the new dataset after imputing the NA values with the mean for its 5-minute interval,

- The mean total number of steps taken each day is `r round(mean_by_day,0)`.  
- The median total number of steps taken each day is `r round(median_by_day,0)`.

We could see that after imputing the NA values, the mean and median of the total number of steps taken each day are almost the same as before we imputing the missing values. Therefore, the impacts from imputing missing data on the estimates are negligible in this case.


## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable "week_part" in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r, echo=TRUE}
activity_data_merge$weekday <- weekdays(as.Date(activity_data_merge$date))

activity_data_merge$week_part <- ifelse(activity_data_merge$weekday %in% c('Saturday','Sunday'), 'weekend','weekday') 
```

Let's see the sample of the data

```{r, echo=TRUE, result="asis"}
set.seed(15)
options(width = 100)
sample_n(activity_data_merge, 10)
```

Look at the average number of steps taken, averaged across all weekdays or weekends 

```{r, echo=TRUE, fig.width=8}
activity_by_interval_and_week_part <- ddply(activity_data_merge, c("week_part", "interval"), summarize, average_steps = mean(steps_modified, na.rm=T))

ggplot(data=activity_by_interval_and_week_part, aes(x=interval, y=average_steps)) + geom_line() +
    facet_grid(week_part~.) + 
    labs(x="5-minute interval", y="average number of steps taken") +
    labs(title="Average number of steps taken for each 5-minute interval across all weekdays or weekends") 
```